SoK: Application Attack Defenses (in Question)

### 2014-10-07 docs/questioning.tex
various additions and better definitions

As in medical diagnostic tests, application attack defense mechanisms
must demonstrate the presence of an attack. This, however, does not
make an application attack defense mechanism immediately useful. In
order for a detection mechanism to be useful it must be accurate, easy to
use, and economical. The accuracy of a detection mechanism is gauged
with the following metrics~\cite{GFDLS06,A00}:
\begin{itemize}
\item {\bf Sensitivity}, the probability that an attack will be
  caught.
\item {\bf Specificity}, the probability that a normal interaction
  will test negative.
\item {\bf Positive Predictive Value} ({\sc ppv}), the probability that a
  reported attack is a real attack.
\item {\bf Negative Predictive Value} ({\sc npv}), the probability that if
  nothing is reported no attack has taken place.
\end{itemize}

For example, a simple announcement that an attack defense mechanism
catches an attack says very little about the usability and
applicability of the mechanism. There may be some value in indicating
that detection \emph{per se} is possible, if that has been considered
infeasible in the past. Most value, though, comes from knowing the
answers to questions like the following, which correspond to the
previous metrics:
\begin{itemize}
\item Does it catch all the attacks? If not, how many of them?
\item Does it catch only attacks? If not, how many false alarms would
  we expect?
\item How much should we be worried if we do get an alarm?
\item How much at ease should we be if no alarm is raised?
\end{itemize}

Sensitivity, specificity, {\sc ppv}, and {\sc npv}, are defined precisely by way
of the $2\times 2$ Table~\ref{tab:attacks-outcomes}. In the cells of
the table we distinguish:
\begin{itemize}
\item True Positive ({\sc tp}), a real attack that raises an alarm.
\item True Negative ({\sc tn}), an event that is not an attack and that does
  not raise an alarm.
\item False Positive ({\sc fp}), an event that although it is not an attack
  raises an alarm.
\item False Negative ({\sc fn}), an event that although is as attack does
  not raise an alarm.
\end{itemize}

@@

\begin{equation}
\textrm{Sensitivity} = \frac{\textrm{{\sc tp}}}{\textrm{{\sc tp}} + \textrm{{\sc tn}}}
\end{equation}

\begin{equation}
\textrm{Specificity} = \frac{\textrm{{\sc tp}}}{\textrm{{\sc tn}} + \textrm{{\sc fp}}}
\end{equation}

\begin{equation}
\textrm{{\sc ppv}} = \frac{\textrm{{\sc tp}}}{\textrm{{\sc tp}} + \textrm{{\sc fp}}}
\end{equation}

\begin{equation}
\textrm{{\sc npv}} = \frac{\textrm{{\sc tn}}}{\textrm{{\sc tn}} + \textrm{{\sc fn}}}
\end{equation}

@@

\caption{Attacks and Test Outcomes}
\label{tab:attacks-outcomes}

\begin{tabular}{l|l|c|c|}
\multicolumn{2}{c}{} & \multicolumn{2}{c}{Attack} \\ \cline{3-4}
\multicolumn{2}{c|}{} & Yes & No \\ \cline{2-4}
\multirow{2}{*}{Result} &  Yes &  True Positive & False Positive \\
& No & False Negative & True Negative \\ \cline{2-4}
\end{tabular}

@@

Sensitivity and specificity can be calculated based on test results
alone. To calculate the sensitivity, we run the test on a controlled
environment where we allow only attack events to reach the system. The
ratio of reported attacks over all events will give us the
sensitivity. Similarly, to calculate the specificity we can run the
test on a controlled environment where we allow only innocuous events
to reach the system. The ratio of non-reported events over all events
will give us the specificity.

The calculation of {\sc ppv} and {\sc npv} requires that we evaluate the attack
defense mechanism on a real operational environment where the events
are a mix of attacks and innocuous events and, moreover, the mix ratio
is the ratio of attacks over harmless events that exists in actual
operations. Too few or too many attacks will skew metrics calculations.

The last requirement may seem cumbersome, but consider that, in actual
operations, {\sc ppv} is what would determine our actions. Administrators
will limited resources need to allocate their attention judiciously.
They may need to prioritize responses to emergency warnings; the {\sc ppv}
value will indicate them what priority to give to a raised alarm.

>>>>

As in medical diagnostic tests, application attack defense mechanisms
must demonstrate the presence of an attack. This, however, does not
make an application attack defense mechanism immediately useful. In
order for a detection mechanism to be useful it must be accurate, easy to
use, and economical. The accuracy of a detection mechanism is gauged
with the following metrics~\cite{TDR2013,GFDLS06,A00}:
\begin{itemize}
\item {\bf Sensitivity}, the probability that an attack will be
  caught.
\item {\bf Specificity}, the probability that a normal interaction
  will test negative.
\item {\bf Positive Predictive Value} ({\sc ppv}), the probability that a
  reported attack is a real attack. It is the conditional probability
  that an event is an attack if the detection mechanism flags it as
  such. 
\item {\bf Negative Predictive Value} ({\sc npv}), the probability that if
  nothing is reported no attack has taken place. It is the conditional
  probability that an event is not an attack given that the detection
  mechanism flags it as normal.
\end{itemize}

Sensitivity and specifity are defined by way of the $2\times 2$
Table~\ref{tab:sensitivity-specificity}~\cite{linn2004}. 
In the cells of
the table we distinguish:
\begin{itemize}
\item True Positive ($a$), a real attack that raises an alarm.
\item True Negative ($d$), an event that is not an attack and that does
  not raise an alarm.
\item False Positive ($b$), an event that although it is not an attack
  raises an alarm.
\item False Negative ($c$), an event that although is as attack does
  not raise an alarm.
\end{itemize}

@@

\begin{equation}
\textrm{SE} = \textrm{Sensitivity} = \frac{a}{a + c}
\end{equation}

\begin{equation}
\textrm{SP} = \textrm{Specificity} = \frac{d}{b + d}
\end{equation}

Sensitivity and specificity can be calculated based on test
data alone. To calculate the sensitivity, we run the test on a
controlled environment where we allow only attack events to reach the
system. The ratio of reported attacks over all attacks will give us
the sensitivity. Similarly, to calculate the specificity we can run
the test on a controlled environment where we allow only innocuous
events to reach the system. The ratio of non-reported events over all
events will give us the specificity.

@@

\caption{Attacks and Test Outcomes in Test Environment}
\label{tab:sensitivity-specificity}

\begin{tabular}{l|l|c|c|}
\multicolumn{2}{c}{} & \multicolumn{2}{c}{Attack} \\ \cline{3-4}
\multicolumn{2}{c|}{} & Yes & No \\ \cline{2-4}
\multirow{2}{*}{Result} &  Yes &  $a = \textrm{True Positive}$ & 
$b = \textrm{False Positive}$ \\
& No & $c = \textrm{False Negative}$ & $d = \textrm{True Negative}$ \\ 
\cline{2-4}
\multicolumn{2}{r|}{Total} & $a + c$ & $b + d$ \\
\cline{3-4}
\end{tabular}

@@

Note that the calculations are carried out vertically; the ratio of
attack events to normal events does not enter into sensitivity and
specificity. This is what allows us to derive them based on test data,
without using data from a real production environment. At the same
time, however, this is what limits their usefulness. When we obtain a
test result in a production environment, be it negative or positive,
we really want to know how much we should be worried. In other words,
if an attack is detected in a production environment, how much should
we be worried? If no attack is detected in a production environment,
how relaxed should we be that no attack has indeed taken place?

\caption{Attacks and Test Outcomes in Real Environment}
\label{tab:ppv-npv}

\begin{tabular}{l|l|c|c|c}
\multicolumn{2}{c}{} & \multicolumn{2}{c}{Attack} & \\ \cline{3-4}
\multicolumn{2}{c|}{} & Yes & No & Total \\ \cline{2-5}
\multirow{2}{*}{Result} &  Yes &  $A = \textrm{True Positive}$ & 
$B = \textrm{False Positive}$ & \multicolumn{1}{c|}{$A + B$}\\
& No & $C = \textrm{False Negative}$ & $D = \textrm{True Negative}$ &
\multicolumn{1}{c|}{$C + D$}\\ 
\cline{2-5}
\end{tabular}

The answer to these questions is given by the \textsc{ppv} and the
\textsc{npv}. These depend not only on the test results, but also on
the \emph{prevalence} of attacks in real environments. In particular,
we can calculate them using Table~\ref{tab:ppv-npv}, which tabulates
results in a real environment, or in an environment where the
prevalence of attacks is equal to what we expect in the real world. To
indicate the difference with Table~\ref{tab:sensitivity-specificity}
we use capital letters for the entries in the table~\cite{linn2004}.
With this table, \textsc{ppv} and \textsc{npv} are given by the
following equations:

\begin{equation}
\textrm{PPV} = \frac{A}{A + C}
\end{equation}

\begin{equation}
\textrm{NPV} = \frac{D}{B + D}
\end{equation}

This time the ratios are taken horizontally; the ratio of attacks to
innocuous events affects the calculations. In fact, if PR is the real
probability that an attack happens, the prevalence, then we
have~\cite{linn2004,altman1994}:

\begin{equation}
\textrm{PPV} = \frac{\textrm{SE}\times \textrm{PR}}{
\textrm{SE}\times \textrm{PR} + (1 - \textrm{SP})\times (1 - \textrm{PR})}
\end{equation}

\begin{equation}
\textrm{NPV} = \frac{\textrm{SP}\times (1 - \textrm{PR})}{
(1 - \textrm{SE})\times \textrm{PR} + \textrm{SP}\times (1 - \textrm{PR})}
\end{equation}

The prevalence is the prior probability that an event is an attack,
based on our understanding of the volume and frequency of attacks; the
\textsc{ppv} and \textsc{npv} are the revised estimates of the
probability based on the results of the detection
mechanism~\cite{altman1994}.
